---
title: "Economic indicators disaggregation"
author: "Sbrana, Sabatella"
date: "2025-12-04"
output:
  html_document:
    theme: simplex
  pdf_document:
    latex_engine: xelatex
editor_options:
  markdown:
    wrap: sentence
---

Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(cowplot)
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(sf)
library(tidyverse)
library(broom)

```

### Data loading and setup

Sets the general working directory where all datasets are stored (wd).

```{r load data, message=FALSE, warning=FALSE, include=FALSE}
wd = "C:/Users/alice/OneDrive - CNR/Documenti - SURIMI-Internal-grp/General/Deliverable 2.1-2.2/" 

```

Defines the GSA codes, country, and supra-region used as filters in the analysis.

```{r CS, message=FALSE, warning=FALSE}
GSAa_CS = c("GSA6")
GSAs_CS = c("GSA06")
Country_CS = "SPAIN"
supra_reg_CS = "MBS"
gear_CS = c("OTB")
```

Loads the GSA shapefile and keeps only the GSA of interest.

```{r GSA, message=FALSE, warning=FALSE}
GSA = read_sf("Data Input/GSAs_simplified.shp") %>%
  filter(SECT_COD == GSAs_CS)
```

Load datasets:

\- **AER**: Economic fleet segment data.

\- **csquare**: Spatial fishing grid cells intersected with the selected GSA.

\- **FDI_effort**: Fishing effort per c-square, filtered so that only cells present in the GSA are retained.

```{r data, message=FALSE, warning=FALSE}
AER <- read.csv(paste0(wd,"Surimi data lake - CNR/Economic and social/AER_economic_fleet segment.csv"))

csquare = read_sf(paste0(wd, "Surimi data lake - CNR/Spatial/EU27/EU27_total_effort_csquares.shp")) %>% 
  st_intersection(GSA)

FDI_effort_sf <- read.csv(paste0(wd, "Surimi data lake - CNR/Spatial/EU27/spatial_effort_tableau_pts_EU27_2012-2024.csv")) %>% 
              filter(gear_type %in% gear_CS) %>% 
                inner_join(csquare) %>% 
                select(year,vessel_length,gear_type, supra_region, cscode,totfishdays, geometry) %>% 
                st_as_sf()
```

Loads a layer of European country borders and reprojects it to match the CRS of the GSA, which is required for correct spatial plotting.
Extracts the bounding box of the GSA to limit the spatial extent of the final maps.

```{r set boundaries, message=FALSE, warning=FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf", continent = "europe")
world = st_transform(world, crs = st_crs(GSA))

xmin = as.numeric(st_bbox(GSA)[1])-0.0001
xmax = as.numeric(st_bbox(GSA)[3])+0.0001
ymin = as.numeric(st_bbox(GSA)[2])-0.0001
ymax = as.numeric(st_bbox(GSA)[4])+0.0001
```

### Standardizing fields

Ensures consistent formatting and whitespace removal for country and supra-region names.

```{r standardize data fields, message=FALSE, warning=FALSE}
## Harmonize country codes
AER$country_name <- toupper(AER$country_name)

## Harmonize supra-region labels (NAO, MBS, OFR)
AER$supra_region <- str_trim(AER$supra_reg)
FDI_effort_sf$supra_region <- str_trim(FDI_effort_sf$supra_region)

```

Filters the AER dataset by year and fishing technique.
Recodes DTS → OTB to harmonize economic and effort datasets.

```{r Compute GVA and Gross Profit from AER, message=FALSE, warning=FALSE}
AER <- AER %>%
  filter(fishing_tech %in% c("DTS")) %>% 
  mutate(gear_type = recode(fishing_tech, 
                            "DTS" = "OTB")) %>% 
  filter(variable_name %in% c("Fishing days","Energy costs", "Energy consumption", "Gross value of landings","Other income",
                              "Repair & maintenance costs", "Other variable costs","Personnel costs",
                              "Value of unpaid labour","Other non-variable costs")) %>%
  mutate(variable_name =  gsub(" ", "_", variable_name)) %>% 
  mutate(variable_name =  gsub("_&", "", variable_name)) %>% 
  mutate(variable_name =  gsub("-", "_", variable_name))

```

Converts the AER table from long to wide format, aggregating economic variables.
Selects only the country and supra-region of interest.


```{r Fuel cost, message=FALSE, warning=FALSE, fig.width= 13, fig.height= 8}
AER_wide <- AER %>%
  dplyr::group_by(country_name, gear_type, year, vessel_length, variable_name) %>%
  dplyr::summarise(total_value = sum(value, na.rm = TRUE)) %>%
  pivot_wider(names_from  = variable_name,
              values_from = total_value)

AER_wide = AER_wide %>% filter(country_name %in% Country_CS)


```


### Join AER - FDI

The second stage of the analysis involves merging these two datasets. The fields that are used to combine the two subdasets AER and FDI are: (1) country_name, (2) year, (3) fishing_tech, (4) vessel_length. We obtained a new dataframe, called aer.fdi.effort, which aligns the values of Total.Fishing.Days from AER and FDI. We obtained a new dataframe, called aer.fdi.effort, in which the values of the Fishing days from AER and FDI are aligned with respect to country_name, year, fishing_tech and vessel_length. In addition, the percentage of the corresponding composition of the Fishing days from FDI by fishing technique is reported. Finally, the ratio between the values of Fishing days from the two data sources is computed. 

A preliminary inspection of the ratio values were performed.

```{r join AER FDI, message=FALSE, warning=FALSE}

FDI_check_fishday = FDI_effort_sf %>%
                    st_drop_geometry() %>% 
                    group_by(year,vessel_length, gear_type) %>%
                    summarise(sum_days = sum(totfishdays))


a = FDI_check_fishday %>% 
  left_join(AER_wide, by = c("year","vessel_length", "gear_type")) %>% 
  mutate(ratio = sum_days/Fishing_days * 100) 

a %>% 
  ggplot(aes(y = Fishing_days, x = sum_days))+
  geom_point()+
  geom_smooth(aes(fill = vessel_length), method = "lm")+
  xlab("FDI fishing days")+
  ylab("AER fishing days")+
  facet_wrap(~ vessel_length)+
  theme_light()

```

### Redistribute economic variables 

#### Applying Sala et al. fuel consumption model

Defines the parameters of the Sala model: Table 8 Sala et al., 2022 dFC = m × LOA + q → daily fuel consumption as a function of vessel length.
(<https://www.nature.com/articles/s41597-022-01478-0>)

Then, it is also possible to observe a series of linear relationships between Fishing Days and the economic parameters (Energy consumption, and Energy costs).
Having the parameters of this relationship, it is possible to use the spatial information in the FDI to transform the spatial values of Fishing Days into corresponding values of Energy consumption and Energy costs. A series of linear models was fitted using these data.
Namely, the Consumption coefficient for each Country, Fishing Technique and Vessel length was estimated using the formula: 

Energy consumption = 𝛽 x Fishing days

The values of 𝛽 obtained using this approach are presented and visually depicted.

```{r Fishing Days and the economic parameters, message=FALSE, warning=FALSE}

AER_wide %>% 
  ggplot(aes(x = Fishing_days, y = Energy_costs, fill = vessel_length))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~ vessel_length, scale = "free")+
  theme_light()+
  ggtitle("Energy consumption = 𝛽 x Fishing days")


coeff_table_FishDay_EC<- AER_wide %>%
group_by(vessel_length) %>%
do(tidy(lm(Energy_costs ~ Fishing_days, data = .))) %>%
ungroup() %>%
filter(term == "Fishing_days") %>% 
select(vessel_length, estimate, p.value) %>%
rename(beta_EC = estimate) %>% 
drop_na()


```

Then, the relationship between Energy Costs and Energy consumption was investigated. Namely, the Cost coefficient for each Country was estimated using the formula:
Energy Costs =  𝛽 x Energy consumption

```{r coeff energy consumption, message=FALSE, warning=FALSE}

AER_wide %>% 
  ggplot(aes(x = Energy_consumption, y = Energy_costs, fill = vessel_length))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~ vessel_length, scale = "free")+
  theme_light()+
  ggtitle("Energy Costs =  𝛽 x Energy consumption")

coeff_table_EC_Econs<- AER_wide %>%
group_by(vessel_length) %>%
do(tidy(lm(Energy_costs ~ Energy_consumption, data = .))) %>%
ungroup() %>%
filter(term == "Energy_consumption") %>% 
select(vessel_length, estimate, p.value) %>%
rename(beta_Econs = estimate) %>% 
drop_na()


```

Add beta coefficient to FDI data

```{r add beta coef to FDI, message=FALSE, warning=FALSE, fig.width= 14, fig.height= 7}
FDI_effort_CS <- FDI_effort_sf %>%
                st_drop_geometry() %>%
                group_by(cscode,year,vessel_length, gear_type, supra_region) %>%
                summarise(totfishdays = round(sum(totfishdays, na.rm = TRUE),1)) %>%
                group_by(year,vessel_length, gear_type,supra_region) %>%
                mutate(sum_days = sum(totfishdays)) %>%
                mutate(effort_share = ifelse(sum_days > 0,
                              totfishdays / sum_days,
                              0))


FDI_effort_CS_beta = FDI_effort_CS %>% 
                      inner_join(coeff_table_FishDay_EC[,1:2]) %>% 
                      inner_join(coeff_table_EC_Econs[,1:2])
  
fdi_aer_disagg = FDI_effort_CS_beta %>% 
                  mutate(Energy_consumption = beta_EC * totfishdays,
                         Energy_Costs = beta_Econs * Energy_consumption)

ggplot()+
  geom_bar(data = fdi_aer_disagg, aes(x = Energy_Costs, y = vessel_length), stat = "identity", position = "dodge")+
  facet_wrap(~ year, nrow = 2)+
  theme_light()+
  ggtitle("Fuel cost by Year")


```


```{r disaggregate by tot fish days, message=FALSE, warning=FALSE}
# 
AER_wide_sel = AER_wide %>%  select(year, vessel_length,gear_type,
                     Gross_value_of_landings,Other_income,
                    Repair_maintenance_costs, Other_variable_costs,Other_non_variable_costs,Personnel_costs,
                             Value_of_unpaid_labour)

aer_fdi <- fdi_aer_disagg %>%
            inner_join(AER_wide_sel, by = c("year" ,"vessel_length", "gear_type"))

aer_fdi_disagg <- aer_fdi %>%
   mutate( GVL_disagg  = Gross_value_of_landings * effort_share,
    Other_income_disagg = Other_income * effort_share,
    Repair_maintenance_costs_disagg = Repair_maintenance_costs*effort_share,
    Other_variable_costs_disagg = Other_variable_costs * effort_share,
    Other_non_variable_costs_disagg = Other_non_variable_costs * effort_share,
    Personnel_costs_disagg =  Personnel_costs * effort_share,
    Value_of_unpaid_labour_disagg = Value_of_unpaid_labour * effort_share) %>% 
   
   select(cscode, year ,vessel_length, gear_type, supra_region, GVL_disagg, Other_income_disagg, Repair_maintenance_costs_disagg,
         Other_variable_costs_disagg, Other_non_variable_costs_disagg, Personnel_costs_disagg, Value_of_unpaid_labour_disagg,
         totfishdays, Energy_consumption,Energy_Costs )

x =   aer_fdi_disagg %>% 
  mutate(
    GVA = GVL_disagg + Other_income_disagg -
      Energy_Costs - Repair_maintenance_costs_disagg -
      Other_variable_costs_disagg - Other_non_variable_costs_disagg,
    
    Gross_profit = GVL_disagg + Other_income_disagg -
      Personnel_costs_disagg - Value_of_unpaid_labour_disagg -
      Energy_Costs - Repair_maintenance_costs_disagg -
      Other_variable_costs_disagg - Other_non_variable_costs_disagg)

```

```{r}
xmap = x %>% 
      group_by(cscode, vessel_length) %>% 
      drop_na() %>% 
      summarise(EC = mean(Energy_Costs),
                GVA = mean(GVA),
                Gross_profit  = mean(Gross_profit),
                totfishdays = mean(totfishdays))

g1 = xmap %>% 
  inner_join(csquare) %>% 
  st_as_sf %>% 
  
  ggplot() +
  geom_sf(data = world, fill = "grey") + 
  geom_sf(aes(fill = EC), colour = NA, inherit.aes = TRUE) + 
  scale_fill_viridis_c("Energy Cost (Euros)", option = "magma", 
                       na.value = NA) + 
  facet_wrap(~ vessel_length)+
  theme_cowplot() +  
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        legend.key.width = unit(0.8, "cm"),     
        legend.key.height = unit(0.4, "cm"),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 5),
        legend.box.margin = margin(t = 5, b = 5),
        legend.box.spacing = unit(0.1, "cm"))+
  scale_x_continuous(limits = c(xmin,xmax)) +
  scale_y_continuous(limits = c(ymin,ymax))+
  ggtitle("Energy Cost")
print(g1)

g2 = xmap %>% 
  inner_join(csquare) %>% 
  st_as_sf %>% 
  ggplot() +
  geom_sf(data = world, fill = "grey") + 
  geom_sf(aes(fill = GVA), colour = NA, inherit.aes = TRUE) + 
  scale_fill_viridis_c("GVA", option = "cividis", 
                      na.value = NA) + 
  theme_cowplot() +  
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        legend.key.width = unit(0.8, "cm"),     
        legend.key.height = unit(0.4, "cm"),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 5),
        legend.box.margin = margin(t = 5, b = 5),
        legend.box.spacing = unit(0.1, "cm"))+
  scale_x_continuous(limits = c(xmin,xmax)) +
  scale_y_continuous(limits = c(ymin,ymax))+
    facet_wrap(~ vessel_length)+
  ggtitle("Gross Value Added")

print(g2)

g3 = xmap %>% 
  inner_join(csquare) %>% 
  st_as_sf %>% 
  ggplot() +
  geom_sf(data = world, fill = "grey") + 
  geom_sf(aes(fill = Gross_profit), colour = NA, inherit.aes = TRUE) + 
  scale_fill_viridis_c("Gross_profit", option = "rocket", 
                        na.value = NA) + 
  theme_cowplot() +  
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        legend.key.width = unit(0.8, "cm"),     
        legend.key.height = unit(0.4, "cm"),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 5),
        legend.box.margin = margin(t = 5, b = 5),
        legend.box.spacing = unit(0.1, "cm"))+
  scale_x_continuous(limits = c(xmin,xmax)) +
  scale_y_continuous(limits = c(ymin,ymax))+
    facet_wrap(~ vessel_length)+
  ggtitle("Gross Profits")

print(g3)

g4 = xmap %>% 
  inner_join(csquare) %>% 
  st_as_sf %>% 
  ggplot() +
  geom_sf(data = world, fill = "grey") + 
  geom_sf(aes(fill = totfishdays), colour = NA, inherit.aes = TRUE) + 
  scale_fill_viridis_c("Fishing days", option = "plasma", 
                        na.value = NA) + 
  theme_cowplot() +  
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        legend.key.width = unit(0.8, "cm"),     
        legend.key.height = unit(0.4, "cm"),
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 5),
        legend.box.margin = margin(t = 5, b = 5),
        legend.box.spacing = unit(0.1, "cm"))+
  scale_x_continuous(limits = c(xmin,xmax)) +
  scale_y_continuous(limits = c(ymin,ymax))+
    facet_wrap(~ vessel_length)+

  ggtitle("Fishing days")

print(g4)


write.csv(x, "Spatial_Economic_value.csv",
          row.names = FALSE)
```




