---
title: "Price disaggregation protocol."
author: "Sbrana, Sabatella"
date: "2025-12-04"
output: 
  html_document:
    theme: united
    toc: true
    toc_float: true
    highlight: tango
  word_document:
    toc: true
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: sentence
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The protocol is part of the SURIMI project and is designed to extract the prices of individual species over time for each landing port.

To implement this methodology effectively, two key datasets are required:

1.  The [FDI landing dataset](https://jeodpp.jrc.ec.europa.eu/ftp/jrc-opendata/FAD/fdi/FDI_spatial_data_EU28.zip), which provides detailed records of landings over time by year, country, GSA, gear, vessel length and ices-cell.

2.  The [FAO Geographical Sub-Areas](https://gfcmsitestorage.blob.core.windows.net/website/5.Data/ArcGIS/GSAs_simplified_updated_division%20(2).zip): Area of application, comprised of the Mediterranean and the Black Sea, as Major Fishing Area 37.

3.  The [FAO ASFIS List of Species for Fisheries](https://www.fao.org/fishery/static/ASFIS/ASFIS_sp.zip) which represents the standard taxonomic reference system for the FAO Statistics Team.

This information is crucial for providing context to the landing data and for connecting fishing effort with species availability and price patterns at the port level.

## Disaggregation procedure

Save the data input to a folder and set the folder as the data location in the R environment:

```{r workdir }
wd = "SET YOUR DATA FOLDER DIRECTORY"

```

```{r wd, include=FALSE}

wd = "C:/Users/alice/OneDrive - CNR/Documenti - SURIMI-Internal-grp/General/Deliverable 2.1-2.2/" 

```

### Data manipulation for a case study area

Users could establish parameters for their case study, which will subsequently inform the procedure. This study examined the activity of bottom otter trawlers (OTB) and purse seiners (PS) within the Spanish waters of GSA 6.

```{r set CS parameters, message=FALSE, warning=FALSE}

CS_name = "FAO GSA06 "
Gear = c("OTB", "PS")
Year = c(2013:2022)
Country = c("ESP")
GSAa_CS = c("GSA6")
GSAs_CS = c("GSA06")

```

```{r library, warning=FALSE,message=FALSE}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(ggpmisc)
library(openxlsx)
library(paletteer)
library(patchwork)
library(reshape2)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(tidyverse)
library(tidytext)
library(webr)
library(webshot2)
```

### Step 1 - Open and subset FDI Landing data

The FDI data were opened and filtered by gear type, year, and country. The GSA polygon was then used to extract spatial landings for Bottom Otter Trawlers (OTB) and Purse Seiners (PS). Figures present selected results showing the total landing values by gear type.

```{r open FDI, eval = TRUE, include=FALSE, cache=TRUE}
GSA = read_sf("Data Input/GSAs_simplified.shp") %>%
       filter(SECT_COD == GSAs_CS)


spatial_landing = read_sf(paste0(wd,"Surimi data lake - CNR/Spatial/FDI_spatial_data_EU28/EU28/landings_csquares.shp")) %>% 
                  st_intersection(GSA)
  


landing = list()

for (y in 1 : length(Year)){

  land_y = read.csv(paste0(wd,"Surimi data lake - CNR/Spatial/FDI_spatial_data_EU28/EU28/spatial_landings_tableau_pts_",
                           Year[y],"_EU28.csv")) %>%
           filter(year %in% Year & gear_type %in% Gear & cscode %in% spatial_landing$cscode) %>%
           mutate(totwghtlandg = as.numeric(totwghtlandg), totvallandg = as.numeric(totvallandg))

  landing[[y]] = land_y
}

names(landing) <- Year



landing_sf = landing_sf <- purrr::map(landing, ~ .x %>%
              left_join(spatial_landing, by = "cscode") %>%
              st_as_sf())


```


```{r plot CS, message=FALSE, warning=FALSE, fig.width=20,fig.height=20, fig.cap= "Total landings coverage for the case study area - resulting from FDI data"}

#Set parameter for the map
world <- ne_countries(scale = "medium", returnclass = "sf", continent = "europe")
world = st_transform(world, crs = st_crs(GSA))

xmin = as.numeric(st_bbox(GSA)[1])-0.0001
xmax = as.numeric(st_bbox(GSA)[3])+0.0001
ymin = as.numeric(st_bbox(GSA)[2])-0.0001
ymax = as.numeric(st_bbox(GSA)[4])+0.0001


#landing

for(g in 1 : length(Gear)){

l_data = purrr::map(landing_sf, ~ .x %>%
           filter(gear_type == Gear[g]))
 
plot_list <- list() 
 
 for(y in 1: length(Year)){
l_plot = l_data[[y]] %>% 
        group_by(year,gear_type,sub_region,cscode) %>% 
        summarise(tot_kg = sum(totwghtlandg), tot_value = sum(totvallandg)) %>% 
       ggplot()+
         geom_sf(aes(fill = tot_value))+
         geom_sf(data = world)+
         coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
         scale_fill_viridis_c(option = "D")+ 
         ggtitle(paste0("FDI EU Landings_",Gear[g],"_",Year[y]))+
         theme_light()

plot_list[[y]] <- l_plot
 }
 combined_plot <- wrap_plots(plotlist = plot_list, ncol = 3)
 print(combined_plot)
 
  cat("-----")
 }

```

Save all the filtered data in a specific folder fd = "CaseStudy/"

```{r save file, message=FALSE, warning=FALSE}

fd = "CaseStudy/"

saveRDS(landing_sf, paste0(fd,"Landing_sf.RData"))

saveRDS(
  purrr::map(landing_sf, ~ .x %>% 
               st_drop_geometry() %>% 
              rename(id = cscode, gear = gear_type, vlength = vessel_length, tot_fish_weight = totwghtlandg,    tot_fish_value = totvallandg)),  
  paste0(fd,"landing_CS.rData"))



```

### Step 2 - Filter and clean landing data

Landing data were processed to identify the key species consistently landed over time and to calculate their average weights, values, and prices. The analysis focuses on catch data by gear type and vessel length (Figure 4) and highlights the top 15 species by landed weight for each gear. Figure 5 presented the time-series of landing values for these selected species, while Figure 6 illustrated the landed kilograms of each species proportionally by gear type.

```{r Filter Landing CS, message=FALSE, warning=FALSE, fig.height= 12, fig.width=8, fig.cap="FDI landing species by Gear (OTB and PS) for the case study area (GSA6)"}
landing_CS = readRDS(paste0(fd,"Landing_CS.RData"))

euro_species = purrr::map(
               landing_CS, ~ .x %>% 
               filter(tot_fish_weight != 0) 
               )

euro_species_df = do.call("rbind", euro_species)

species_all_years <- euro_species_df %>%
  group_by(gear,species) %>%  
  distinct(year, species) %>%
  summarise(n_years = n_distinct(year), .groups = "drop") %>%
  filter(n_years == length(unique(euro_species_df$year)))

euro_species_filtered <- euro_species_df %>%
                         inner_join(species_all_years, by = c("species", "gear"))

# remove quarter

euro_species_filtered <- euro_species_filtered %>%  
                           group_by(year, gear, species) %>% 
                           summarise(yearly_kg = mean(tot_fish_weight)*1000, yearly_value = mean(tot_fish_value))
                           
euro_species_filtered %>%
   group_by(species, gear) %>%
   summarise(mvalue = mean(yearly_value), kg = (mean(yearly_kg))) %>%
  ggplot() +
  geom_bar(aes(x = kg, y = reorder_within(species, kg, gear), fill = gear), stat = "identity")+
  ylab("FAO species code")+
  facet_wrap(~ gear, scales = "free")+
  xlab("kg")+
  ggtitle("List of all landed species for the CS")+
  scale_y_reordered()+
  theme_minimal()
           
ggsave(filename = "Result/land_spe.jpeg", height = 12)

 species_sel <- euro_species_filtered %>%
  group_by(species, gear) %>%
  summarise(m_kg = mean(yearly_kg), .groups = "drop") %>%
  group_by(gear) %>%
  slice_max(m_kg, n = 15) %>%
  ungroup()  

```

### Step 3 - Add species price

```{r results, message=FALSE, warning=FALSE, fig.width=15,fig.height=8}
 
FDI_land_spe_filter <- euro_species_filtered %>%
                    inner_join(species_sel, by = c("species", "gear")) %>% 
                    select(-m_kg) %>% 
                    mutate(price = yearly_value/yearly_kg)

FAO_sp = read.csv("Data Input/ASFIS_sp_2025.csv") %>% select("Alpha3_Code","Scientific_Name") %>% rename("species" = "Alpha3_Code")

FDI_land_spe_filter$year = as.character(FDI_land_spe_filter$year)

gear_filter = unique(FDI_land_spe_filter$gear)

for (g in seq_along(gear_filter)) {

  print(FDI_land_spe_filter %>% 
    filter(gear == gear_filter[g]) %>% 
  group_by(year, species) %>% 
  summarise(m_price = mean(price)) %>% 
  left_join(FAO_sp) %>%
  mutate(sp_id = paste0(Scientific_Name," (",species,")")) %>% 
  
        ggplot(aes(x = year, y = m_price, color = species,
                   group = species)) +
        geom_line(size = 1.1) +
        geom_point() +
        facet_wrap(~ sp_id, scales = "free", ncol = 5) +
        theme_minimal() +
        ggtitle(paste0("Species price time series for ", gear_filter[g])) +
        ylab("Species price (€)")+
        theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
)
}


FDI_land_spe_filter %>%
   group_by(gear,species) %>%
   summarise(kg_gear = sum(yearly_kg)) %>%
  group_by(species) %>% 
  mutate(kg_tot = sum(kg_gear)) %>% 
  mutate(kg_prop = (kg_gear/kg_tot)*100) %>%
  left_join(FAO_sp) %>%
  mutate(sp_id = paste0(Scientific_Name," (",species,")")) %>% 
  ggplot() +
  geom_bar(aes(y = sp_id, x = kg_prop, fill = gear),stat = "identity") +
  labs(title = "Kg of landing by gear", y = "", x = "Kg prop") +
  theme_minimal()+
  theme(axis.text.y = element_text(face = "italic"))

```

Save data filtered

```{r select species, message=FALSE, warning=FALSE, fig.width=8, fig.height=10}

saveRDS(FDI_land_spe_filter, paste0(fd,"FDI_land_spe_filter.RData"))

write.csv(FDI_land_spe_filter, "Result/Species_Prices.csv", row.names = F)

```



